On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-26 paper/results2d.tex
KMK edits

\section{Two-dimensional disks with beta cooling}\label{result_2d}
We begin in the 2D limit with standard beta cooling  to make connection 
with previous studies. The disk
material is assumed to be confined to the midplane and $\delta v_z=0$. 
We make the replacement  
$\rho \to \Sigma$, re-interpret $P$ as the vertically-integrated
pressure, and set $\Gamma=1$. % without loss of generality. 
%big gamma only controls vertical structure
%can absorb big gamma into def of cs
The gravitational potential perturbation remains 3D and its midplane
value is given by    
\begin{align}
  \delta \Phi(z=0) = -\frac{2 \pi G}{|k|}\delta\Sigma
\end{align}
\citep{shu70}.

@@

\subsection{Inviscid limit}\label{2d_inviscid}
We first simplify the problem by setting $\alpha = \alpha_b = 0$. This
eliminates viscous heating and forces, allowing us to quantify the 
\emph{sole} effect of cooling on the 
perturbations. A time-independent heat source should be
invoked to balance the imposed cooling to allow an equilibrium to be
defined ($\mathcal{H}_\mathrm{ext}\neq 0$). Alternatively, we are
assuming viscosity only provides a background heating and does not
play an active role in the perturbed state. %may or may not be true in a real
                                %disk because we are using viscosity
                                %as proxy for turbulence
                                %maybe turb heating actually behaves
                                %like background heat source only

Eq. \ref{thindisk} becomes 
\begin{align}\label{inviscid}
%  &\beta S^3 + S^2 - \beta \left\{ \frac{2|K|}{\qtwo} - \left[2(2-q) +
%    \gamma K^2\right]\right\}S \notag\\ &- \left[\frac{2|K|}{\qtwo} - 2(2-q)\right]
%  = 0. 
  S^2 = \frac{2|K|}{Q} - 2(2-q) - \left(\frac{\theta + \beta \gamma
    S}{1+\beta S}\right)K^2, 
\end{align}
similar to the classic dispersion relation
(Eq. \ref{classic_dr}), which may be obtained by taking the limit
$|\beta S|\to\infty$. %then setting gamma=1. other params fixed
                      %(finite theta)
The first term on the right-hand-side represents destabilization by self-gravity; 
the second and third terms represent stabilization by rotation and
pressure, respectively. The imposed cooling/irradiation only affects the
pressure response. {\bf do we need to explain the terms again?}

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}\label{stable_condition}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. 
%gamma/theta condition is related to some kind of thermal instability
%oscilitory convection?? 
%second condition concerning Q is associated with self-g 
Notice these conditions do not actually
depend on the cooling time. 
The disk becomes more unstable with decreasing $\theta$, 
i.e. as the floor temperature, towards which perturbations are cooled, drops.

If only real growth rates are considered, then violating the second
condition in Eq. \ref{stable_condition} alone is sufficient for instability. 
%from property of cubic crossing the real axis 
In that case the wavenumbers satisfying
\begin{align}
  \theta K^2 - \frac{2}{Q}|K| + 2(2-q) < 0 
\end{align}
are unstable. The range of unstable wavenumbers
increases with decreasing irradiation $\theta$. For $\theta\ll 1$ this range is 
$(2-q)Q\lesssim|K|\lesssim2/\theta Q$. Without irradiation there is no
upper limit to unstable wavenumbers, which could have important
implications for numerical simulations (see \S\ref{prev_works}).

Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p|K| =
0$ and $S = S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
\begin{align}\label{kstar}
%  |K_*| = \frac{1}{\gamma \qtwo} \left(1 + \frac{1}{\beta S_*}\right). 
  |K_*| = \frac{1+\beta S_*}{Q\left(\theta + \gamma \beta S_*\right)}
\end{align}
%Eq. \ref{kstar} describes how the most unstable  mode is affected by
%cooling ($\beta$) and external heating ($\theta$). 
Inserting this into Eq. \ref{inviscid}, we find the maximum growth
rate satisfies
\begin{align}\label{inviscid_max}
%  S_*^3 + \left[2(2-q)  - \frac{1}{\gamma \qtwo^2}\right]S_* - \frac{1}{\gamma
%    \qtwo^2 \beta} = 0.
  S_*^2 = \frac{1+\beta S_*}{Q^2\left(\theta + \beta\gamma S_*\right)}
  - 2(2-q).
\end{align}
Eq. \ref{kstar}---\ref{inviscid_max} imply $\p_\beta|K_*|,\,\p_\beta
S_*<0 $ for $\gamma>\theta$. That is, as cooling becomes more rapid
the disk becomes more unstable at shorter wavelengths. 
%show dK*/dbeta < 0 
%by diff K* wrt beta and diff S* wrt beta from the equations
%and elim dS*/d beta

@@

\subsubsection{$\theta = 0$}\label{theta0}
For beta cooling with $\theta=0$, the disk is unconditionally unstable
for finite $Q$.  The $\beta\to0$ limit corresponds to a 
pressureless disk (not merely isothermal).

%for definiteness 
Let us consider a disk with 
\begin{align}\label{qcrit_def}
  \qtwo = \frac{1}{\sqrt{2\gamma(2-q)}}\equiv Q_{\mathrm{crit}},
\end{align} 
so that it is marginally stable in the absence of cooling.  
How does finite cooling destabilize the disk?  
Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} and setting
$\theta=0$, we find 
\begin{align}\label{sstar}
  S_*^3 = \frac{1}{\gamma Q_\mathrm{crit}^2 \beta}. 
\end{align}
The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
increases with decreasing $\beta$. Instability formally occurs for
arbitrarily long but finite cooling times. 
%Instability does not require the
%cooling time to be below some critical value.

@@

\begin{align}\label{betastar}
  \beta_* = \frac{1}{\left(\sqrt{\gamma} - 1\right)^{3/2}}. 
\end{align}
This equation reproduce similar values 
of the cooling times below which 
%dynamical disk 
%fragmentation occurs in numerical simulations
numerical simulations show dynamical disk fragmentation 
\citep{gammie01,rice05,rice11}. These simulations employ the same beta  
cooling prescription with $\theta=0$, and determine the fragmentation
boundary, $\beta_c$, as a function of the adiabatic index $\gamma$.  
Table \ref{bstar_compare} shows rough agreement between $\beta_*$ and 
$\beta_c$.

@@

\subsubsection{$\theta = 1$}\label{theta1}
Standard beta cooling with $\theta=1$ corresponds to 
`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case 
$\beta\to 0$ corresponds to an isothermal disk and $\beta
\to \infty$ corresponds to an adiabatic disk. (These two limits are
equivalent for $\gamma=1$.) The most unstable wavenumber and growth
rates vary weakly with $\beta$.

@@

To see the effect of cooling and irradiation, we simplify the
dispersion relation, Eq. \ref{thindisk}, by assuming $|\beta S|\ll
1$. Then for $|K| \to 0$ we find
\begin{align}\label{gammie_smallk}
%  S \sim \frac{\alpha |K|^3}{(2-q)Q},%derived assuming |beta*S|<<1 
%  S \sim \frac{\alpha K^2\left[2|K|Q^{-1} - \gamma K^2 + 2\alpha q
%    (2-q)(\gamma-1)K^2\right]}{2(2-q)+\gamma K^2 - 2|K|Q^{-1}} %derived
                                %assuming |\beta S|>>1
  S\simeq \frac{\alpha K^2}{2(2-q)}\left(\frac{2|K|}{Q} - \theta
  K^2\right), 
\end{align}
which coincides with \citeauthor{gammie96}'s Eq. 18 for vanishing
wavenumber. For $|K|\to\infty$ we find
\begin{align}\label{gammie_bigk}
  %  S \sim \frac{2}{Q|K|}\left(\frac{4}{3}\alpha + \gamma\beta\right)^{-1}.
  S \simeq\left(\frac{2}{Q|K|} - \theta\right)\left(\frac{4}{3}\alpha + 
  \alpha_b + \gamma\beta\right)^{-1}.
\end{align}

@@

As a numerical example, we consider a model with $\alpha_b=0$ and
$\alpha=\alpha(\beta)$ given by thermal equilibrium (
Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
\begin{align}
  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
\end{align}
%with $Q_\mathrm{crit}$ given by 
to mimic a basic state maintained by gravito-turbulence where one
might expect the dimensionless stress $\alpha \sim Q^{-2}$
\citep{lin87}.   
%should really only use this Q-alpha relation for order-unity values
%of Q

Fig. \ref{gammie_rate_plot} show growth rates as a function of the
wavenumber $k$ obtained from the dispersion relation
Eq. \ref{thindisk}. The limiting behavior for small/large $K$ are
well-captured by Eq. \ref{gammie_smallk} and
\ref{gammie_bigk}. Comparing the two panels shows that increasing the
irradiation level ($\theta$) surpresses small-scale
perturbations. %(This also occurs in the inviscid case.)

@@

Fig. \ref{gammie_maxrate_plot} shows the maximum
growth rate (top panel) and the corresponding wavenumber (bottom
panel) as a function of the cooling time $\beta$ for $\theta=0$.  
There is good agreement between numerical growth rates and
Eq. \ref{gammie_maxrate_simple} for $\beta \gtrsim 1$. 
%The assumption
%of $|\beta S|\ll 1$ is not well satisfied at smaller $\beta$ which
%explains the mismatch there. 
Eq. \ref{gammie_maxrate_simple} gives the
limiting behavior for this case as   
\begin{align*}
  S_*\propto \begin{cases}
%    \frac{\alpha^{1/4}}{Q\beta^{3/4}} \propto \beta^{-3/2} 
    \alpha^{1/4}\beta^{-3/4}Q^{-1}\propto \beta^{-3/2} 
    &  \beta \gg \alpha, \\
%    \frac{1}{Q\sqrt{\alpha}}  
    \alpha^{-1/2}Q^{-1} = \mathrm{const.} & \beta \ll \alpha,    
  \end{cases}
\end{align*}
where we have applied Eq. \ref{alpha_beta_relation}
and \ref{Qalpha}. The disk is formally unstable for all $\beta$, but
growth timescales are long ($>10$ orbits) for $\beta \gtrsim
20$. %increase beta -> drop alpha -> increase Q. latter two expected
     %to stablize
This region is marked by the vertical dashed-dotted line in
Fig. \ref{gammie_maxrate_plot}. The optimum wavenumber decreases with
the cooling time for $\beta\lesssim O(1)$ because larger scales are
more resistant to the associated increase in viscous damping.

%{\bf maybe move following to discussion}
Numerical simulations of gravito-turbulent disks show there is a
maximum $\alpha$ ($\sim 0.06$) that can be sustained before
fragmentation \citep{rice11}. We can interpret this result in our
linear framework. 
%We can use our case (\ref{gvisc}) to interpret this
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$. Fig. \ref{gammie_maxrate_plot} 
shows that such a system would be dynamically unstable with 
growth rate $s = O(\Omega)$. This is due to the direct effect of viscous 
stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 
and hence highly turbulent, self-gravitating disks to persist beyond
dynamical timescales. This is consistent with previous numerical
simulations performed by \cite{lodato05}. 
%Case (\ref{gvisc}) is consistent with 
\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    1.0cm]{figures/result2d_gvisc_kmax}
  \caption{Growth rates (top) at the optimal wavenumber (bottom) for
    viscous GI as a function of cooling time for the
    case shown in Fig. \ref{gammie_rate_plot}. Solid lines are
    computed from the dispersion relation (Eq. \ref{thindisk}), and the dotted line is an
    estimate from Eq. \ref{gammie_maxrate_simple}. The vertical dashed
    line marks the region with $\alpha > 1$, and the vertical
    dashed-dot line marks the region with growth timescales longer
    than dynamical. 
    \label{gammie_maxrate_plot}}
\end{figure}

>>>>

\section{Two-dimensional disks with beta cooling}\label{result_2d}
We begin in the 2D limit with standard beta cooling  to facilitate comparison 
with previous studies. The disk
material is assumed to be confined to the midplane, and $\delta v_z=0$. 
We make the replacement  
$\rho \to \Sigma$, re-interpret $P$ as the vertically-integrated
pressure, and set $\Gamma=1$. % without loss of generality. 
%big gamma only controls vertical structure
%can absorb big gamma into def of cs
The gravitational potential perturbation remains 3D and its midplane
value is given by    
\begin{align}
  \delta \Phi(z=0) = -\frac{2 \pi G}{|k|}\delta\Sigma
\end{align}
\citep{shu70}.

@@

\subsection{Inviscid limit}\label{2d_inviscid}
We first simplify the problem by setting $\alpha = \alpha_b = 0$. This
eliminates viscous heating and forces, allowing us to quantify the 
\emph{sole} effect of cooling on the 
perturbations. %KMK
We emphasize that destabilization is independent of the effect of decreasing temperature on the instantaneous
value of the classic-$Q$. A time-independent heat source should be
invoked to balance the imposed cooling to allow an equilibrium to be
defined ($\mathcal{H}_\mathrm{ext}\neq 0$). Alternatively, we are
assuming viscosity only provides a background heating and does not
play an active role in the perturbed state. %may or may not be true in a real
                                %disk because we are using viscosity
                                %as proxy for turbulence
                                %maybe turb heating actually behaves
                                %like background heat source only 
	%KMK interesting points. discuss later?
Eq. \ref{thindisk} becomes 
\begin{align}\label{inviscid}
%  &\beta S^3 + S^2 - \beta \left\{ \frac{2|K|}{\qtwo} - \left[2(2-q) +
%    \gamma K^2\right]\right\}S \notag\\ &- \left[\frac{2|K|}{\qtwo} - 2(2-q)\right]
%  = 0. 
  S^2 = \frac{2|K|}{Q} - 2(2-q) - \left(\frac{\theta + \beta \gamma
    S}{1+\beta S}\right)K^2, 
\end{align}
similar to the classic dispersion relation
(Eq. \ref{classic_dr}), which may be obtained by taking the limit
$|\beta S|\to\infty$. %then setting gamma=1. other params fixed
                      %(finite theta)
The first term on the right-hand-side represents destabilization by self-gravity; 
the second and third terms represent stabilization by rotation and
pressure, respectively. The imposed cooling/irradiation only affects the
pressure response. %KMK no...? {\bf do we need to explain the terms again?}

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}\label{stable_condition}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. 
%gamma/theta condition is related to some kind of thermal instability
%oscilitory convection?? 
%second condition concerning Q is associated with self-g 
Notice these conditions do not actually
depend on the cooling time. 
The disk becomes more unstable with decreasing $\theta$, 
i.e. as the floor temperature, towards which perturbations are cooled, drops. 
{\bf KMK maybe more words here about apparent trend of $Q \rightarrow \inf$ for small $\theta$
even forward ref. lengthscales below?}

If only real growth rates are considered, then violating the second
condition in Eq. \ref{stable_condition} alone is sufficient for instability. 
%from property of cubic crossing the real axis 
In that case the wavenumbers satisfying
\begin{align}
  \theta K^2 - \frac{2}{Q}|K| + 2(2-q) < 0 
\end{align}
are unstable. The range of unstable wavenumbers
increases with decreasing irradiation $\theta$. For $\theta\ll 1$ this range is 
$(2-q)Q\lesssim|K|\lesssim2/\theta Q$. Without irradiation there is no
upper limit to unstable wavenumbers, which could have important
implications for numerical simulations probing large wavenumbers and small scales
at high resolution (see \S\ref{prev_works}).

Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p|K| =
0$ and $S = S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
\begin{align}\label{kstar}
%  |K_*| = \frac{1}{\gamma \qtwo} \left(1 + \frac{1}{\beta S_*}\right). 
  |K_*| = \frac{1+\beta S_*}{Q\left(\theta + \gamma \beta S_*\right)}
\end{align}
%Eq. \ref{kstar} describes how the most unstable  mode is affected by
%cooling ($\beta$) and external heating ($\theta$). 
Inserting this into Eq. \ref{inviscid}, we find the maximum growth
rate satisfies
\begin{align}\label{inviscid_max}
%  S_*^3 + \left[2(2-q)  - \frac{1}{\gamma \qtwo^2}\right]S_* - \frac{1}{\gamma
%    \qtwo^2 \beta} = 0.
  S_*^2 = \frac{1+\beta S_*}{Q^2\left(\theta + \beta\gamma S_*\right)}
  - 2(2-q).
\end{align}
Eq. \ref{kstar}---\ref{inviscid_max} imply $\p_\beta|K_*|,\,\p_\beta
S_*<0 $ for $\gamma>\theta$. That is, as cooling becomes more rapid
the disk becomes more unstable at shorter wavelengths.  {\bf KMK -- I don't follow}
%show dK*/dbeta < 0 
%by diff K* wrt beta and diff S* wrt beta from the equations
%and elim dS*/d beta

@@

\subsubsection{$\theta = 0$}\label{theta0}
For beta cooling with $\theta=0$, the disk is unconditionally unstable
for finite $Q$. {\bf KMK some caveating about scales}  The $\beta\to0$ limit corresponds to a 
pressureless disk (not merely isothermal).

%for definiteness 
Let us consider a disk with 
\begin{align}\label{qcrit_def}
  \qtwo = \frac{1}{\sqrt{2\gamma(2-q)}}\equiv Q_{\mathrm{crit}},
\end{align} 
{\bf KMK $\gamma = \theta$ from eq. 47? I'm missing something}
so that it is marginally stable in the absence of cooling.  
How does finite cooling destabilize the disk?  
Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} and setting
$\theta=0$, we find 
\begin{align}\label{sstar}
  S_*^3 = \frac{1}{\gamma Q_\mathrm{crit}^2 \beta}. 
\end{align}
The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
increases with decreasing $\beta$. Instability formally occurs for
arbitrarily long but finite cooling times. 
%Instability does not require the
%cooling time to be below some critical value.

@@

\begin{align}\label{betastar}
  \beta_* = \frac{1}{\left(\sqrt{\gamma} - 1\right)^{3/2}}. 
\end{align}
This equation reproduces remarkably similar values 
of the cooling times below which 
%dynamical disk 
%fragmentation occurs in numerical simulations
numerical simulations show dynamical disk fragmentation 
\citep{gammie01,rice05,rice11}. These simulations employ the same beta  
cooling prescription with $\theta=0$, and determine the fragmentation
boundary, $\beta_c$, as a function of the adiabatic index $\gamma$.  
Table \ref{bstar_compare} shows rough agreement between $\beta_*$ and 
$\beta_c$.

@@

\subsubsection{$\theta = 1$}\label{theta1}
Standard beta cooling with $\theta=1$ corresponds to 
`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case 
$\beta\to 0$ corresponds to an isothermal disk and $\beta
\to \infty$ corresponds to an adiabatic disk. (These two limits are
equivalent for $\gamma=1$.) The most unstable wavenumber and growth
rates vary weakly with $\beta$.   {\bf KMK: summarize boring result? not sure 
what this simplifies to in comparison to other limits?}.

@@

%KMK moved fig ref. 
Fig. \ref{gammie_rate_plot} shows growth rates as a function of the
wavenumber $k$ obtained from the dispersion relation
Eq. \ref{thindisk}.  To compute growth rates numerically, 
 we consider a model with $\alpha_b=0$ and
$\alpha=\alpha(\beta)$ given by thermal equilibrium (
Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
\begin{align}
  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
\end{align}
%with $Q_\mathrm{crit}$ given by 
to mimic a basic state maintained by gravito-turbulence where one
might expect the dimensionless stress $\alpha \sim Q^{-2}$
\citep{lin87}.

It is useful to consider several limiting cases.
To see the effect of cooling and irradiation, we simplify the
dispersion relation, Eq. \ref{thindisk}, by assuming $|\beta S|\ll
1$. Then for $|K| \to 0$ we find
\begin{align}\label{gammie_smallk}
%  S \sim \frac{\alpha |K|^3}{(2-q)Q},%derived assuming |beta*S|<<1 
%  S \sim \frac{\alpha K^2\left[2|K|Q^{-1} - \gamma K^2 + 2\alpha q
%    (2-q)(\gamma-1)K^2\right]}{2(2-q)+\gamma K^2 - 2|K|Q^{-1}} %derived
                                %assuming |\beta S|>>1
  S\simeq \frac{\alpha K^2}{2(2-q)}\left(\frac{2|K|}{Q} - \theta
  K^2\right), 
\end{align}
which coincides with \citeauthor{gammie96}'s Eq. 18 for vanishing
wavenumber. For $|K|\to\infty$ we find
\begin{align}\label{gammie_bigk}
  %  S \sim \frac{2}{Q|K|}\left(\frac{4}{3}\alpha + \gamma\beta\right)^{-1}.
  S \simeq\left(\frac{2}{Q|K|} - \theta\right)\left(\frac{4}{3}\alpha + 
  \alpha_b + \gamma\beta\right)^{-1}.
\end{align}

@@

%%KMK MOVED?
%As a numerical example, we consider a model with $\alpha_b=0$ and
%$\alpha=\alpha(\beta)$ given by thermal equilibrium (
%Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
%\begin{align}
 % Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
%\end{align}
%%with $Q_\mathrm{crit}$ given by 
%to mimic a basic state maintained by gravito-turbulence where one
%might expect the dimensionless stress $\alpha \sim Q^{-2}$
%\citep{lin87}.   
%should really only use this Q-alpha relation for order-unity values
%of Q

%Fig. \ref{gammie_rate_plot} shows growth rates as a function of the
%wavenumber $k$ obtained from the dispersion relation
%Eq. \ref{thindisk}. 
The limiting behavior for small/large $K$ are
well-captured by Eq. \ref{gammie_smallk} and
\ref{gammie_bigk}. Comparing the two panels of Fig. \ref{gammie_rate_plots} shows that increasing the
irradiation level ($\theta$) supresses small-scale
perturbations. %(This also occurs in the inviscid case.)

@@

Fig. \ref{gammie_maxrate_plot} shows the maximum
growth rate (top panel) and the corresponding wavenumber (bottom
panel) as a function of the cooling time $\beta$ for $\theta=0$.  
There is good agreement between numerical growth rates and
Eq. \ref{gammie_maxrate_simple} for $\beta \gtrsim 1$. 
%The assumption
%of $|\beta S|\ll 1$ is not well satisfied at smaller $\beta$ which
%explains the mismatch there. 
Eq. \ref{gammie_maxrate_simple} gives the
limiting behavior for this case as   
\begin{align*}
  S_*\propto \begin{cases}
%    \frac{\alpha^{1/4}}{Q\beta^{3/4}} \propto \beta^{-3/2} 
    \alpha^{1/4}\beta^{-3/4}Q^{-1}\propto \beta^{-3/2} 
    &  \beta \gg \alpha, \\
%    \frac{1}{Q\sqrt{\alpha}}  
    \alpha^{-1/2}Q^{-1} = \mathrm{const.} & \beta \ll \alpha,    
  \end{cases}
\end{align*}
{\bf KMK physically what is the limit $\beta \ll \alpha$?}
where we have applied Eq. \ref{alpha_beta_relation}
and \ref{Qalpha}. The disk is formally unstable for all $\beta$, but
growth timescales are long ($>10$ orbits) for $\beta \gtrsim
20$. %increase beta -> drop alpha -> increase Q. latter two expected
     %to stablize
This region is marked by the vertical dashed-dotted line in
Fig. \ref{gammie_maxrate_plot}. The optimum wavenumber decreases with
the cooling time for $\beta\lesssim O(1)$ because larger scales are
more resistant to the associated increase in viscous damping.   {\bf KMK this is
confusing b/c viscosity is in general acting on larger scales than pressure. also damping -> dissipation or 
angular momentum transport?}

%{\bf maybe move following to discussion}
Numerical simulations of gravito-turbulent disks show there is a
maximum $\alpha$ ($\sim 0.06$) that can be sustained before
fragmentation \citep{rice11} {\bf KMK rice 05a,b, no?}. We can interpret this result in our
linear framework. 
%We can use our case (\ref{gvisc}) to interpret this
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$. Fig. \ref{gammie_maxrate_plot} 
shows that such a system would be dynamically unstable with 
growth rate $s = O(\Omega)$. This is due to the direct effect of viscous 
stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 
and hence highly turbulent, self-gravitating disks to persist beyond
dynamical timescales.
%KMK
We interpret fragmentation as the response to the instability of the
highly viscous state. This is consistent with previous numerical
simulations performed by \cite{lodato05}. 
%Case (\ref{gvisc}) is consistent with 
\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    1.0cm]{figures/result2d_gvisc_kmax}
  \caption{Growth rates (top) at the optimal wavenumber (bottom) for
    viscous GI as a function of cooling time for the
    case shown in Fig. \ref{gammie_rate_plot}. Solid lines are
    computed from the dispersion relation (Eq. \ref{thindisk}), and the dotted line is an
    estimate from Eq. \ref{gammie_maxrate_simple}. The vertical dashed
    line marks the region with $\alpha > 1$, and the vertical
    dashed-dot line marks the region with growth timescales longer
    than dynamical. 
    \label{gammie_maxrate_plot}}
\end{figure}

### 2016-01-26 paper/results3d.tex
updating figs, need to correct 3d beta results, need to re-comp bg if Q depends on alpha
KMK edits

\section{Three-dimensional disks with beta cooling}\label{3ddisk}
We confirm the above results in 3D disks with vertical
structure. This is expected to weaken gravitational instabilities
because the disk mass is spread across the vertical dimension. It is
possible to incorportate this effect in the previous 2D framework, but this
introduces additional parameters as discussed in Appendix
\ref{3dcorr}. It is more direct to solve the 3D eigenvalue problem to
avoid such uncertainties. Our numerical approach is outlined in Appendix
\ref{3d_method}.

@@

In the following examples we consider a vertically isothermal disk
($\Gamma=1$). Then in the viscous case $\alpha$ is vertically
constant (Eq. \ref{alpha_beta_relation}). We consider even modes about
the midplane, and apply a numerical disk surface at $z=\zmax$ such
that $\rho(\zmax)=0.05\rho_0$.

@@

We consider a 3D inviscid disk with $\qthree=0.71$ and $\theta=0$. 
The 3D gravity parameter $\qthree$ is defined by 
Eq. \ref{Q3d_def}. For this case the corresponding Toomre parameter 
$Q=2Q_\mathrm{crit}$. Recall $Q_\mathrm{crit}$, defined by
Eq. \ref{qcrit_def}, is the Toomre parameter value such that the 2D
disk would be marginally stable in the absence of cooling.

Fig. \ref{3d_inviscid} show growth rates and the most unstable
wavenumber obtained for this model. We also plot 2D
results with the 3D correction as described in Appendix
\ref{3dcorr}. The (empirically) chosen value of
$H_\mathrm{sg}=0.64H$ results in a close match between 2D and 3D
growth rates, but the most unstable wavenumber in 3D is somewhat smaller. 
This reflects self-gravity being weakened in the vertical 
direction: a larger horizontal scale is required to achieve the same
strength of self-gravity as the 2D case. Similarly, 
choosing $H_\mathrm{sg}=0.53H$ matches the optimum wavenumbers, but
growth rates are over-estimated in 2D.

@@

Fig. \ref{3d_visc} show growth rates, maximized over $k$, as a
function of the cooling time $\beta$. We also plot 2D results with 3D
corrections. Softening the self-gravity in 2D captures the correct 
qualitative behavior as in the full 3D case, but it is clear that a
single, constant value of $H_\mathrm{sg}$ cannot re-produce 3D growth 
rates for all $\beta$. 
This suggests that the exact value of
$H_\mathrm{sg}$ is problem-dependent, although taking
$H_\mathrm{sg}\sim O(H)$ should give the correct 3D growth rate within
a factor of two.

%cooling helps vertical modes (convection) 
%slower variation wrt z for low beta because of associated high
%viscosity
%smooth out vertical gradients 
Fig. \ref{3d_visc_vz} shows the magnitude of the vertical velocity
$|\delta v_z|$ scaled by the total horizontal velocity for $\beta =
1,\,10$ and $100$. Cooling generally enhances vertical motion
(convection). However, even in the highly unstable case with $\beta=1$,
vertical speeds are sub-dominant at $\sim 30\%$ of the horizontal
speeds. This suggests that it is not essential to consider 3D models
when considering gravitational instabilities.

>>>>

\section{Three-dimensional disks with beta cooling}\label{3ddisk}
We confirm the above results in 3D disks with vertical
structure. Accounting for the third dimension will weaken gravitational instabilities
because the disk mass is spread across some vertical extent. It is
possible to incorporate this effect in the previous 2D framework, but doing so
introduces additional parameters as discussed in Appendix
\ref{3dcorr}. It is more direct to solve the 3D eigenvalue problem to
avoid such uncertainties. Our numerical approach is outlined in Appendix
\ref{3d_method}.

@@

In the following examples we consider a vertically isothermal disk
($\Gamma=1$). Then in the viscous case $\alpha$ is vertically
constant (Eq. \ref{alpha_beta_relation}). We consider {\bf only} even modes about
the midplane, and apply a numerical disk surface at $z=\zmax$ such
that $\rho(\zmax)=0.05\rho_0$.

@@

We consider a 3D inviscid disk with $\qthree=0.71$ and $\theta=0$. 
The 3D gravity parameter $\qthree$ is defined by 
Eq. \ref{Q3d_def}. For such a disk, the corresponding Toomre parameter 
$Q=2Q_\mathrm{crit}$. Recall $Q_\mathrm{crit}$, defined by
Eq. \ref{qcrit_def}, is the Toomre parameter value such that the 2D
disk would be marginally stable in the absence of cooling.

Fig. \ref{3d_inviscid} shows growth rates and the most unstable
wavenumbers obtained for this model. We also plot 2D
results with the 3D correction as described in Appendix
\ref{3dcorr}. The (empirically) chosen value of
$H_\mathrm{sg}=0.64H$, also defined in Appendix \ref{3dcorr}, results in a close match between 2D and 3D
growth rates, but the most unstable wavenumber in 3D is somewhat smaller. 
This offset reflects self-gravity being weakened in the vertical 
direction: a larger horizontal scale is required to achieve the same
strength of self-gravity as the 2D case. Similarly, 
choosing $H_\mathrm{sg}=0.53H$ matches the optimum wavenumbers, but
growth rates are over-estimated in 2D.

@@

Fig. \ref{3d_visc} shows growth rates, maximized over $k$, as a
function of the cooling time $\beta$. We also plot 2D results with 3D
corrections. Softening the self-gravity in 2D captures the correct 
qualitative behavior as in the full 3D case, but once again it is clear that a
single, constant value of $H_\mathrm{sg}$ cannot re-produce 3D growth 
rates for all $\beta$. 
This suggests that the exact value of
$H_\mathrm{sg}$ is problem-dependent, although taking
$H_\mathrm{sg}\sim O(H)$ should give the correct 3D growth rate within
a factor of two.

%cooling helps vertical modes (convection) 
%slower variation wrt z for low beta because of associated high
%viscosity
%smooth out vertical gradients 
Fig. \ref{3d_visc_vz} shows the magnitude of the vertical velocity
$|\delta v_z|$ scaled by the total horizontal velocity for $\beta =
1,\,10$ and $100$. Cooling generally enhances vertical motion
(convection) {\bf KMK really convection?}. However, even in the highly unstable case with $\beta=1$,
vertical speeds are sub-dominant at $\sim 30\%$ of the horizontal
{\bf KMK  azimuthal or radial?} speeds. This suggests that it is not essential to consider 3D models
when considering gravitational instabilities. {\bf KMK this is not true generically, so
this statement needs to be more specific or cut}.
